openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2020-10-01 paper/policar2020-opentsne.tex
revision of the main text

Exaggeration can highlight transitions between cell states in developmental studies. Standard t-SNE often produces embeddings with clearly defined, discrete clusters. We can adjust the level of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are often undesired in developmental studies where cells' state is changed continuously. To this end, other embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} tend to capture the continuity between cell states better. Recently, B{\"o}hm \textit{et al.}~\cite{bohm2020unifying} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ construct embeddings which are markedly similar to UMAP and ForceAtlas2, respectively. For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time points is difficult to observe due to many sprawled out clusters. On the other hand, it is easier to trace the development when we increase the exaggeration factor from $1$ to $4$ (Fig.~\ref{fig:cao}.b-c).

@@

t-distributed stochastic neighbor embedding (t-SNE) is a non-linear
dimensionality reduction method, which aims to find a low-dimensional
embedding, where local neighborhoods are preserved. More formally, given a
multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2,
\dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the number of data
points in the data set, t-SNE aims to find a low dimensional embedding
$\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right
\} \in \mathbb{R}^d$ where $d \ll D$, such that if points $\mathbf{x}_i$ and
$\mathbf{x}_j$ are close in the high-dimensional space, their corresponding
embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is
primarily used as a visualization tool, $d$ is typically set to two. The
similarity between two data points in the high-dimensional space is defined as

@@

\noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying
density of the data in the multi-dimensional space. Perplexity is sometimes
interpreted as the continuous analog to the number of nearest neighbors to which
distances will attempt to be preserved.

The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the
embedding space is defined using the $t$-distribution with a single degree of
freedom (Cauchy kernel)

@@

The Kullback-Leibler (KL) divergence is used as a measure of agreement
between distributions $\mathbf{P}$ and $\mathbf{Q}$

@@

\noindent The objective is to find embeddings $\mathbf{Y}$ such that the
KL divergence is minimized.

The corresponding gradient takes the form
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) w_{ij},
\label{eq:tsne_gradient}
\end{equation}
where $w_{ij} = \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}$
and represents the unnormalized $q_{ij}$.

Optimization is performed with batch gradient descent using the delta-bar-delta
update rule~\cite{jacobs1988increased}. Originally, t-SNE was run for 1000
iterations consisting of two phases: in the first \textit{early exaggeration}
phase, the attractive forces between data points are increased by some factor
$\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the space and find
their respective neighbors. The remaining 750 iterations are run with $\rho=1$,
which reverts the attractive forces to their original values and produces the
final embedding.

Belkina \textit{et al.} later found that faster convergence can be achieved by
increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. As a side-effect, embeddings converge faster, and the
number of iterations can be lowered to 750, decreasing the overall runtime.
This convention has been adopted by most modern t-SNE implementations.

\subsection*{Efficient Approximation Schemes} A direct evaluation of t-SNE
gradients requires $\mathcal{O}(N^2)$ operations, which makes its application
impractical to any reasonably-sized data set and beckons for the development of
efficient approximation schemes. Van der Maaten observed that the t-SNE
gradient may be cast as an N-body problem where data points represent
particles which attract and repel each
other~\cite{van2014accelerating}. The gradient from Eqn. (\ref{eq:tsne_gradient}) can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \left [ \sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) -\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \right ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2
\right )^{-1}$. This equation can then be viewed as a particle simulation, where 
the two terms represent the attractive and repulsive
forces between individual particles. Each term lends itself to efficient
approximations, which enable us to greatly reduce the time complexity of t-SNE.

@@

Van der Maaten observed that evaluating the attractive forces between all pairs
of data points is excessive, and that considering only a handful of nearest
neighbors at each point is sufficient to obtain a good approximation~\cite{van2014accelerating}.
Because distances are transformed to similarities using a Gaussian kernel, and
the bandwidth of each kernel is selected such that only a predefined number of
neighbors fall within the main probability mass of the bell curve through the perplexity parameter, the remaining
data points fall into the exponentially decaying tails of each Gaussian distribution.
These data points are assigned near-zero probabilities and do not contribute
to the overall attractive forces of data points.
Therefore it is sufficient to calculate the attractive forces for only a
small number of nearest neighbors instead of all $N$ points. By
utilizing tree-based nearest-neighbor search methods, the time complexity is
thus reduced to $\mathcal{O}(N \log N)$. Linderman \textit{et al.} further
realized that, qualitatively, embeddings are visually
indistinguishable when using only \textit{approximate} nearest neighbors,
further reducing time complexity to $\mathcal{O}(N)$~\cite{linderman2019fast}.

@@

Examining the second term of Eqn.~(\ref{eq:grad_attr_rep}) we notice that each
point indiscriminately exerts a repulsive force on all other points. Van der
Maaten proposed an approach based on N-body simulations and used a
space-partitioning Barnes-Hut tree approach to approximate the interaction
between data points~\cite{van2014accelerating}. Briefly, in the 2D case, the approach splits
the space into quadrants and entire regions may be summarized by simple statistics.
If a query point is far away from a given quadrant, the repulsive forces exerted
by all the points in that quadrant onto the query point are summarized by a single point. This
reduces the time complexity from $\mathcal{O}(N^2)$ to
$\mathcal{O}(N \log N)$.

More recently, Linderman \textit{et al.} proposed an alternative approach,
FIt-SNE, based on non-uniform convolutions for calculating all pairwise
interactions between repelling data points~\cite{linderman2019fast}. Briefly,
Linderman \textit{et al.} observed that the repulsive forces $\mathbf{R}$ from Eqn. (\ref{eq:grad_attr_rep}) may
be rewritten as \begin{align}
\mathbf{R}_i &= \sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \notag \\
&= \sum_{j \neq i} \frac{\mathbf{y}_i - \mathbf{y}_j}{\left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 \right )^{2}}
\bigg/
\sum_{k \neq l} \frac{1}{1 + || \mathbf{y}_k - \mathbf{y}_l || ^2} \label{eq:fitsne_rep}
\end{align} 
and computed by evaluating three terms
\begin{align}
\phi_{1, j} &= \sum_{j \neq i} \frac{1}{1 + || \mathbf{y}_j - \mathbf{y}_i ||^2}, \notag \\
\phi_{2, j} &= \sum_{j \neq i} \frac{\mathbf{y}_j}{\left( 1 + || \mathbf{y}_j - \mathbf{y}_i ||^2 \right)^2}, \notag \\
\phi_{3, j} &= \sum_{j \neq i} \frac{1}{\left( 1 + || \mathbf{y}_j - \mathbf{y}_i ||^2 \right)^2}. \label{eq:fitsne_terms}
\end{align}
Then the numerator and denominator of Eqn.~(\ref{eq:fitsne_rep}) can be
computed as $\mathbf{y}_i \phi_{1,j} - \phi_{2,j}$ and $Z = \sum_j \phi_{3,j}$,
respectively. These interactions are calculated by interpolating the terms
through a grid of equispaced interpolation points. This shifts the
computational burden onto the interpolation points and reduces the time
complexity to $\mathcal{O}(N)$.

% -------- pointless...
%\opentsne\ provides efficient implementations for all the aforementioned
%improvements but defaults to using approximate nearest neighbor search and the
%non-uniform convolutions approach as these enable the user to quickly create
%informative visualizations for data sets containing up to millions of samples.

@@

t-SNE is non-parametric and does not define an explicit mapping from the
high-dimensional space to the embedding space. Therefore embeddings of new data
points need to be found through the use of optimization
techniques~\cite{policar2019embedding}. When adding new data points to an
existing, reference embedding, the reference data points are fixed in place while
new data points are allowed to find their respective positions. The
optimization remains the same as in standard t-SNE with only slight
modifications to $p_{ij}$ and $q_{ij}$
\begin{align}
p_{j \mid i} &= \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) /  \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \\
q_{j \mid i} &= \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}},
\end{align}
\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Eqn.~(\ref{eq:symmetrize}).
Plugging these terms into Eqn.~(\ref{eq:kl_divergence}), we obtain the following
gradient
\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}.
\label{eq:gradient}
\end{equation}

Similarly to standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N
\cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE
approximation schemes, reducing the time complexity to $\mathcal{O}(M \log N)$
and $\mathcal{O}(\max \{ N, M \})$, respectively. In the FIt-SNE approximation
scheme, we additionally exploit the fact that the reference embedding remains
fixed throughout the optimization of newly added points, and precompute the
interpolation grid. This further reduces the runtime complexity from
$\mathcal{O}(\max \{ N, M \})$ to $\mathcal{O}(M)$.

@@

In standard t-SNE, distances are converted to similarities through the use of
Gaussian kernels of varying bandwidths. The bandwidths are indirectly
determined by the user-specified perplexity parameter so that a fixed number of
nearest data points will be assigned non-zero values. One common trick for
uncovering the global relations between clusters is to increase perplexity so
that more long-range interactions are preserved in the final embedding.
However, one unfortunate side effect of increasing perplexity is that smaller
clusters get absorbed into larger ones.

Kobak \& Berens suggest that replacing the Gaussian kernel
with a mixture of Gaussians may provide better insight into both the local and
global structure~\cite{kobak2019art}. For instance, the similarities between data points in the input
space may instead be computed with

@@

The bandwidth of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined
by selecting different perplexity values. Using a kernel with
perplexity 500 captures long-range interactions which preserve global
cluster organization. Combining these with a more narrow kernel with
perplexity 50 prevents small, well-defined clusters from being absorbed
into larger ones, leading a better overall snapshot into the data structure.

@@

Standard t-SNE reveals the clustering structure at a single level of
resolution. Different perplexity parameter values can be used to identify
global cluster relationships or small, well-isolated groups. Unfortunately,
varying perplexity values can be time-consuming as this involves recomputing
the $k$-nearest neighbor graph which is often the most expensive part of the t-SNE algorithm.
Alternatively, Kobak \textit{et al.} suggest that varying
the degree of freedom in the t-distribution can be used to explore the clustering structure at different
levels of resolution~\cite{kobak2019heavy}.

Standard t-SNE models similarities between data points in the embedding space
using a t-distribution with a single degree of freedom, but this can be generalized
to allow for any parameter value

\begin{equation}
q_{ij} \propto \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right )^{-\alpha} = \frac{1}{\left( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right)^\alpha }.
\end{equation}
In standard t-SNE $\alpha=1$ so this simplifies to the Cauchy kernel from
Eqn.~(\ref{eq:cauchy_kernel}). The gradient of the loss function then becomes
\begin{align}
\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) w_{ij}^{1/\alpha} \left ( \mathbf{y}_i - \mathbf{y}_j \right ),
\end{align}
which can, again, be cast as the interplay between the attractive and repulsive forces between particles
\begin{align}
\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} p_{ij} w_{ij}^{1/\alpha} (\mathbf{y}_i - \mathbf{y}_j) \notag \\
&- 4 \sum_{j \neq i} w_{ij}^{\frac{\alpha+1}{\alpha}} / Z (\mathbf{y}_i - \mathbf{y}_j).
\end{align}

Adapting existing approximation schemes to this formulation is straightforward.
\opentsne\ provides efficient implementations of both the Barnes-Hut and the FIt-SNE
approximation schemes, where we modify the terms from
Eqn.~(\ref{eq:fitsne_terms}) to

@@

Embeddings produced by standard t-SNE often produce clusters separated by thin
boundaries and make use of all available space. While this is desirable at times,
it often obscures the global relationships between clusters as all
neighboring clusters appear about the same distance from one another. Other
dimensionality reduction methods such as UMAP~\cite{2018arXivUMAP} or
ForceAtlas2~\cite{jacomy2014forceatlas2} tend to produce embeddings where
clusters appear more compact and the white-space separating the clusters may be
interpreted at least partially as a loose measure of distance.

B{\"o}hm \textit{et al.} showed that the exaggeration factor
$\rho$ can be used to produce layouts more similar to UMAP and ForceAtlas2~\cite{bohm2020unifying}.
By incorporating exaggeration into later phases of optimization, t-SNE is able
introduce more white-space between clusters which better reflects the global
relations between clusters. B{\"o}hm \textit{et al.} found that using $\rho=4$ and $\rho=30$
produces embeddings visually similar to UMAP and ForceAtlas2, respectively.

While it is difficult to claim one is better than the other, different settings
of $\rho$ may uncover different properties of the data manifold. For
example, in single-cell data, standard t-SNE often uncovers
distinct, well-separated groups of cell-types but obscures transitional paths between cell-states. This problem is
exacerbated when dealing with large numbers of data points. On the other hand,
ForceAtlas2 has been used successfuly to uncover trajectories and transitions between
cell types. Unfortunately, when using ForceAtlas2, large clusters often absorb small, distinct groups of cells.
One might run t-SNE with various parameter settings to avoid switching between
different algorithms and obtain similar visualizations.
We might generate an embedding using higher levels of exaggeration to highlight
the developmental transitions between cell types and
one with low levels of exaggeration to identify clear populations of cells.

@@

t-SNE, UMAP, and ForceAtlas2 can all be cast as force-directed layout algorithms,
which operate on the $k$-nearest neighbor graph. Each method constructs the graph differently
and specifies the attractive and repulsive forces in its own way, but ultimately,
point positions are found by balancing the attractive and repulsive forces between data points.

The final positions of data points in this class of algorithms is largely
dependent on the embedding initialization. UMAP found early success as it was
able to produce visualizations that better captured the global organization of
clusters. This was due to its initialization scheme, which initialized point
positions using Laplacian eigenmaps. On the other hand, most implementations of
t-SNE performed random initialization, which resulted in poor global coherence.
However, Kobak \& Linderman recently showed that using the same initialization
for both methods leads to visualizations, which exhibit similarly good global
coherency. Conversely, when initialized randomly, both methods produce
visualizations where clusters are arbitrarily positioned in the embedding space.

\opentsne\ defaults to using the two leading principal components as
initialization, but also provides a spectral approach similar to UMAP.
This allows clusters in resulting embeddings to be organized in a more globally coherent
manner, leading to increased visualization interpretability.

@@

\opentsne\ is distributed under the BSD-3-Clause License and is publicly
available as an open-source package at
\url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available on
\texttt{PyPI} and \texttt{conda-forge}. The data sets used and scripts used in this study are
included in accompanying notebooks, publicly available at
\url{https://github.com/pavlin-policar/opentsne-paper}.

>>>>

Exaggeration can highlight transitions between cell states in developmental studies. Standard t-SNE often produces embeddings with clearly defined, discrete clusters. We can adjust the level of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are often undesired in developmental studies where cells' state is changed continuously. To this end, other embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} tend to capture the continuity between cell states better. Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ construct embeddings which are markedly similar to UMAP and ForceAtlas2, respectively. For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time points is difficult to observe due to many sprawled out clusters. On the other hand, it is easier to trace the development when we increase the exaggeration factor from $1$ to $4$ (Fig.~\ref{fig:cao}.b-c).

@@

t-distributed stochastic neighbor embedding (t-SNE) is a non-linear
dimensionality reduction method, which aims to find a low-dimensional
embedding, where local neighborhoods are preserved. More formally, given a
multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2,
\dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the number of data
points in the data set, t-SNE aims to find a low dimensional embedding
$\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right
\} \in \mathbb{R}^d$ where $d \ll D$, such that if points $\mathbf{x}_i$ and
$\mathbf{x}_j$ are close in the high-dimensional space, their corresponding
embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is
primarily used as a visualization tool, $d$ is typically set to two. The
similarity between two data points in the high-dimensional space is defined as:

@@

\noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying
density of the data in the multi-dimensional space. Perplexity is sometimes
considered the continuous analog to the $k$ nearest neighbors to which
distances will be preserved.

The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the
embedding space is defined using the $t$-distribution with a single degree of
freedom. This is also known as the Cauchy kernel:

@@

Finally, the Kullback-Leibler (KL) divergence is used as a measure of agreement
between distributions $\mathbf{P}$ and $\mathbf{Q}$

@@

\noindent The t-SNE objective is to find embeddings $\mathbf{Y}$ such that the
KL divergence is minimized.

The corresponding gradient takes the form
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}.
\label{eq:tsne_gradient}
\end{equation}

Optimization is performed using batch gradient descent with the delta-bar-delta
update rule~\cite{jacobs1988increased}. Originally, t-SNE was run for 1000
iterations consisting of two phases. In the first \textit{early exaggeration}
phase, the attractive forces between data points is increased by a factor
$\rho=12$ so that points can more easily move throughout the embedding to find
their respective neighbors. The remaining 750 iterations are run with $\rho=1$,
which reverts the attractive forces to their original values and produces the
final embedding.

Belkina \textit{et al.} later found that faster convergence can be achieved by
increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. The
number of iterations can then safely be lowered to 750, leading to faster
runtime. This convention has been adopted by most modern t-SNE implementations.

\subsection*{Efficient Approximation Schemes} A direct evaluation of t-SNE
gradients requires $\mathcal{O}(N^2)$ operations, which makes its application
impractical to any reasonably-sized data set and beckons for the development of
efficient approximation schemes. Van der Maaten observed that the t-SNE
gradient may be reformulated in terms of a N-body problem where data points may
be interpreted as particles which attract and repel each
other~\cite{van2014accelerating}. The gradient may be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \left [ \sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) -\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \right ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2
\right )^{-1}$. The two terms may be viewed as the attractive and repulsive
forces between particles, respectively. Each term lends itself to efficient
approximations, which enable us to greatly reduce the time complexity of t-SNE.

@@

Van der Maaten observed that since $p_{ij}$ is calculated in the
high-dimensional space using a Gaussian kernel and that the bandwidth is
selected such that only a fixed number of closest neighbors (determined by the
perplexity parameter) will contribute to the attractive forces of a given point
due to the exponential decay of the Gaussian kernel~\cite{van2014accelerating}.
Therefore it is sufficient to calculate the attractive forces only over a
relatively small number of nearest neighbors instead of all $N$ points. By
utilizing tree-based nearest-neighbor search methods, the time complexity is
thus reduced to $\mathcal{O}(N \log N)$. Linderman \textit{et al.} further
realized that, qualitatively, embeddings are visually virtually
indistinguishable when using only \textit{approximate} nearest neighbors,
further reducing time complexity to $\mathcal{O}(N)$~\cite{linderman2019fast}.

@@

Examining the second term of Eqn.~(\ref{eq:grad_attr_rep}) we notice that each
point indiscriminately exerts a repulsive force on all other points. Van der
Maaten proposed an approach based on N-body simulations and used a
space-partitioning Barnes-Hut tree approach to approximate the interaction
between data points~\cite{van2014accelerating}. Briefly, the approach splits
the space into quadrants (in the 2D case) and if a given point is far enough
away from a quadrant, the points in that quadrant are summarized by a single
center of mass. This reduces the time complexity from $\mathcal{O}(N^2)$ to
$\mathcal{O}(N \log N)$.

More recently, Linderman \textit{et al.} proposed an alternative approach,
FIt-SNE, based on non-uniform convolutions to calculating all pairwise
interactions between repelling data points~\cite{linderman2019fast}. Briefly,
Linderman \textit{et al.} observed that the repulsive forces $\mathbf{R}$ may
be rewritten as \begin{align}
\mathbf{R}_i &= \sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \notag \\
&= \sum_{j \neq i} \frac{\mathbf{y}_i - \mathbf{y}_j}{\left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 \right )^{2}}
\bigg/
\sum_{k \neq l} \frac{1}{1 + || \mathbf{y}_k - \mathbf{y}_l || ^2} \label{eq:fitsne_rep}
\end{align} 
and computed by evaluating three terms
\begin{align}
\phi_{1, j} &= \sum_{j \neq i} \frac{1}{1 + || \mathbf{y}_j - \mathbf{y}_i ||^2}, \notag \\
\phi_{2, j} &= \sum_{j \neq i} \frac{\mathbf{y}_j}{\left( 1 + || \mathbf{y}_j - \mathbf{y}_i ||^2 \right)^2}, \notag \\
\phi_{3, j} &= \sum_{j \neq i} \frac{1}{\left( 1 + || \mathbf{y}_j - \mathbf{y}_i ||^2 \right)^2}. \label{eq:fitsne_terms}
\end{align}
Then the numerator and denominator of Eqn.~(\ref{eq:fitsne_rep}) can be
computed as $\mathbf{y}_i \phi_{1,j} - \phi_{2,j}$ and $Z = \sum_j \phi_{3,j}$,
respectively. These interactions are accelerated by interpolating these terms
through a grid of equispaced interpolation points. This shifts the
computational burden onto the interpolation points and reduces the time
complexity to $\mathcal{O}(N)$.

\opentsne\ provides efficient implementations for all the aforementioned
improvements but defaults to using approximate nearest neighbor search and the
non-uniform convolutions approach as these enable the user to quickly create
informative visualizations for data sets containing up to millions of samples.

@@

t-SNE is non-parametric and does not define an explicit mapping from the
high-dimensional space to the embedding space. Therefore embeddings of new data
points need to be found through the use of optimization
techniques~\cite{policar2019embedding}. When adding new data points to an
existing, reference embedding, the reference data points are fixed in place and
the new data points are allowed to find their respective positions. The
optimization remains the same as in standard t-SNE with only slight
modifications to $p_{ij}$ and $q_{ij}$
\begin{align}
p_{j \mid i} &= \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) /  \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \\
q_{j \mid i} &= \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}},
\end{align}
\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the secondary data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Eqn.~(\ref{eq:symmetrize}). The gradients of
$\mathbf{w}_j$ with respect to the loss (Eqn.~\ref{eq:kl_divergence}) are:
\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}
\label{eq:gradient}
\end{equation}

As in standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N
\cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE
approximation schemes, reducing the time complexity to $\mathcal{O}(M \log N)$
and $\mathcal{O}(\max \{ N, M \})$, respectively. In the FIt-SNE approximation
scheme, we additionally exploit the fact that the reference embedding remains
fixed throughout the optimization of newly added points, and precompute the
interpolation grid. This further reduces the runtime complexity from
$\mathcal{O}(\max \{ N, M \})$ to $\mathcal{O}(M)$.

@@

In standard t-SNE, distances are converted to similarities through the use of
Gaussian kernels of varying bandwidths. The bandwidths are indirectly
determined by the user-specified perplexity parameter so that a fixed number of
nearest data points will have non-zero similarity. One common trick for
uncovering the global relations between clusters is to increase perplexity so
that more long-range interactions are preserved in the final embedding.
However, one unfortunate side effect of increasing perplexity is that smaller
clusters get absorbed into larger ones.

Kobak \& Berens~\cite{kobak2019art} suggest that replacing the Gaussian kernel
with a mixture of Gaussians may provide better insight into both the local and
global structure. Therefore, the similarities between data points in the input
space become

@@

The bandwidths of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined
by selecting different perplexity values. They show that using a kernel with
perplexity 500 captures the long-range interactions which preserve global
cluster organization in the final embedding and combine this with a kernel with
perplexity 50, which prevents small, well-defined clusters from being absorbed
into larger ones.

@@

Standard t-SNE reveals the clustering structure in data at one level of
resolution. Different perplexity parameter values can be used to identify
global cluster relationships or small, well-isolated groups. Unfortunately,
varying perplexity values can be time-consuming as this involves recomputing
the KNNG which is often the most expensive part of the t-SNE algorithm.
Alternatively, Kobak \textit{et al.}~\cite{kobak2019heavy} suggest that varying
the degree of freedom in the t-distribution used in the embedding space during
optimization can be used to explore the clustering structure at different
levels of resolution.

Standard t-SNE models similarities between data points in the embedding space
using a t-distribution with a single degree of freedom

\begin{equation}
q_{ij} \propto \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right )^{-\alpha} = \frac{1}{\left( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right)^\alpha }.
\end{equation}

In standard t-SNE $\alpha=1$ so this simplifies to the Cauchy kernel from
Eqn.~(\ref{eq:cauchy_kernel}).

The gradient of the loss function then becomes
\begin{align}
\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) q_{ij}^{1/\alpha} \left ( \mathbf{y}_i - \mathbf{y}_j \right ),
\end{align}
which can, again, be decomposed into the attractive and repulsive forces as
\begin{align}
\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} p_{ij} q_{ij}^{1/\alpha} (\mathbf{y}_i - \mathbf{y}_j) \notag \\
&- 4 \sum_{j \neq i} q_{ij}^{\frac{\alpha+1}{\alpha}} / Z (\mathbf{y}_i - \mathbf{y}_j).
\end{align}

Adapting existing approximation schemes to this formulation is straightforward.
We provide efficient implementations of both the Barnes-Hut and the FIt-SNE
approximation schemes, where we replace the terms from
Eqn.~(\ref{eq:fitsne_terms}) with

@@

Embeddings produced by standard t-SNE often produce clusters separated by thin
boundaries and make use of all available space. While this is often desirable,
this often obscures the global relationships between clusters as all
neighboring clusters appear at the same distance from one another. Other
dimensionality reduction methods such as UMAP~\cite{2018arXivUMAP} or
ForceAtlas2~\cite{jacomy2014forceatlas2} tend to produce embeddings where
clusters appear more compact and the white-space separating the clusters may be
interpreted at least partially as a loose measure of distance.

Kobak \textit{et al.}~\cite{todo} showed that the early exaggeration factor
$\rho$ in t-SNE may be increased to produce more compact layouts often seen in
UMAP or ForceAtlas2. Early exaggeration is a trick introduced by van der
Maaten~\cite{maaten2008visualizing} used in the first 250 iterations and
increases the attractive forces between data points, enabling neighboring
points to move close to one another while ignoring the repulsive forces between
data points. In standard t-SNE, the remainder of the iterations are run with
$\rho=1$. Kobak \textit{et al.} showed that using $\rho=4$ and $\rho=30$ in the
second phase of the optimization results in embeddings visually similar to UMAP
embeddings, and ForceAtlas2 embeddings, respectively.

While it is difficult to claim one is better than the other, different settings
of $\rho$ may uncover different properties and clustering structures. For
example, in single-cell data, standard t-SNE with $\rho=1$ often uncovers
distinct groups of cell-types but obscures developmental paths. This problem is
exacerbated when dealing with large numbers of data points. Conversely,
ForceAtlas2 has often been used to uncover trajectories and transitions between
cell types, but larger clusters often absorb small, distinct groups of cells.

@@

Similarly to UMAP and ForceAtlas2, t-SNE can also be viewed as a force-directed
layout algorithm. The first phase of the t-SNE algorithm constructs a
$k$-nearest neighbor graph, and the subsequent optimization balances the
attractive and repulsive forces between data points.

The final positioning of data points using this class of algorithms is largely
dependent on the initial positions in the embedding space. UMAP is often said
to better preserve global structure than t-SNE, but Kobak \& Linderman have
recently discovered that this supposed advantage is entirely due to the better
default initialization in UMAP than in older implementations of
t-SNE~\cite{kobak2019umap}. Standard implementations of t-SNE use random
initializations, which often scatter resulting clusters throughout the
embedding space. UMAP injects global organization into the embedding by
initializing point positions using Laplacian eigenmaps. When initialized in the
same manner, both t-SNE and UMAP perform equally well, or equally poorly,
dispelling the myth that UMAP is superior in this respect.

\opentsne\ defaults to using the two leading principal components as
initialization, but also provides a spectral approach, similar to that of UMAP.
The clusters in resulting embeddings are organized in a globally coherent
manner, leading to more interpretable visualizations.

@@

openTSNE is distributed under the BSD-3-Clause License and is publicly
available as an open-source package at
\url{https://github.com/pavlin-policar/openTSNE}. openTSNE is also available on
PyPI and conda-forge. The data sets used and scripts used in this study are
included in accompanying notebooks, publicly available at
\url{https://github.com/pavlin-policar/opentsne-paper}.
